Compositional Quantitative Resource Analysis of C Programs

### 2014-06-04 paper/main.tex
new title
alternative title (just a comment)
new introduction

\documentclass[10pt]{article}
\usepackage{fullpage}

@@

\title{Automatic Amortized Analysis for Imperative Code}
\date{}
\author{}

\begin{document}
\maketitle

>>>>

\documentclass[nocopyrightspace,preprint]{sigplanconf}

@@

\begin{document}

\conferenceinfo{POPL} {January XX-XX, 2014, CITY, STATE, India.}
\copyrightyear{2014}
\copyrightdata{XXX-X-XXXX-XXXX-X/XX/XX}

\titlebanner{Draft -- Do not distribute}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

% \title{Automatic Amortized Analysis for Imperative Code}
\title{Compositional Quantitative Resource Analysis of C Programs}
% Compositional Resource-Bound Certification for C Programs

\authorinfo
{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
{Yale University}

\maketitle

\begin{abstract}

\end{abstract}

\category{D.2.4}{Software Engineering}
{Software/Program Verification}
\category{F.3.1}{Logics and Meanings of Programs}
{Specifying and Verifying and Reasoning about Programs}

\terms Verification, Reliability

\keywords Formal Verification, Compiler Construction, Program Logics,
Stack-Space Bounds, Quantitative Verification

\section{Introduction}
\label{sec:intro}

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems [XXX (see PLDI)].
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing [XXX?  (cite this guy form the
us))].  Worst-case resource bounds can be used to prevent side-channel
attacks through resource padding [XXX?].  Loop and
recursion-depth bounds are used to ensure the accuracy of programs
that are executed on unreliable hardware [XXX oopsla] and complexity
bounds are needed to verify security protocols [XXX].
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include Speed [XXX], KoAT
[XXX], Pubs [XXX], RANK [XXX], and LOOPUS [XXX].
%
While the these tools can derive impressive results for
realistic software, there still exist several shortcomings that seem
to hamper the application of the existing techniques in practice.
\begin{enumerate}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant generation [XXX speed] or
  translation of the program into a term-rewriting system [costa,
  tacas] without providing \emph{verifiable certificates} for the
  correctness of the derived bound.
\end{enumerate}
%
While there has been progress there is still truth in the following
quote of Don Knuth from a recent interview [xxx].

\begin{quotation}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
\end{quotation}
%
%
In this paper, we present a novel cost analysis system for C programs
that works automatically, efficiently and precisely for many examples,
can be used interactively if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{itemize}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement that is parametric in a
  cost metric.  The semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost
  model.
\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional programs
  [XXX] and derives bounds that are non-negative linear combinations
  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that the our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{itemize}
%
Our approach complements existing works since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource bound development
in which an automatically derived bound can be used in the
quantitative logic and a hand derived bound can be used by the
automatic analysis.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters.  To the best of our knowledge, our
system is the only one that can deal with resource like memory that
may become available during the execution.

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that
seem to be common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
$B(100,200,43) = 1043$.  
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce multiple results from the literature which were obtained
using methods based on abstract interpretation.  In contrast with
abstract interpretation--based methods, our technique does not require
any fixpoint computations on an abstract domain to obtain loop
invariants.  The mechanism we designed is able to leverage local
assertions such as $x < y$ that we collect along the branching points
to obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, mutually-recursive functions.

Following the development steps for automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived though).  More specifically, bounds have the form
$\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer variables
or constants.  The reason is mainly clarity of presentation, since
an extension to multivariate resource polynomials [XXX] would make the
inference rules considerably more complected and should better be
described separately.  However, we developed the linear inference
system so that the extension to polynomial bounds should work
smoothly.

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. XXX

With the intention, of making the material easily accessible we first
informally describe the new automatic amortized analysis
(\pref{sec:inform}).  We then formalize the intuition with the
operational cost semantics an the inference rules for the amortized
analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
quantitative Hoare logic and its soundness proof that is formalized in
Coq.  In \pref{sec:inter} we prove the soundness of the automatic
amortized analysis by showing that derivation using the inference
rules can be seen as derivations in the quantitative logic.  We also
show how to integrate hand-derived bounds with automatically-derived
bounds.  \pref{sec:exper} contains the results of the experimental
evaluation and \pref{sec:related} describes related research.
